Process-based phenotypes: an application to the Cellular Phenotype Ontology

### 2011-11-22 trunk/paper/drinking.tex

In computer science, an ontology is a specification of a
conceptualization of a domain of knowledge \cite{Gruber1995,
  Guarino1998}.  % In philosophy, on the other hand, an ontology is a
% system of categories that account for a particular view on the
% world. The philosophical sense of ``ontology'' is what computer
% science calls ``conceptualization''.  While a conceptualization is
% language-independent, its specification is based on a language. In
% this sense, ontologies specify some aspects of the meaning of terms
% in vocabularies.
Ontologies commonly distinguish between {\em classes} and {\em
  individuals} within a domain of knowledge. A class is an entity that
can have {\em instances}, while individuals are entities that cannot
be instantiated. Examples of individuals include the Eiffel tower or
the 2009 Ironman Triathlon in Hawaii, while examples of classes
include {\em Tower} or {\em Triathlon}. The Eiffel tower can be an
instance of the class {\em Tower}, and the 2009 Ironman Triathlon an
instance of {\em Triathlon}.  The meaning of classes is specified by
stating what must be true of their instances.

@@

A first kind of abnormality of regulatory processes are {\em abnormal
  temporal distribution patterns} of a process. In these
abnormalities, the {\em way} in which processes of a particular kind
are temporally distributed is abnormal.  The most common abnormal
distribution pattern is an increased or decreased frequency, and we
use PATO's {\em frequency} class to define {\em Abnormal frequency of
  occurrence of $X$}.
% as a subclass of an {\em Abnormality of the regulation of $X$}.

>>>>

In computer science, an ontology is a specification of a
conceptualization of a domain of knowledge \cite{Gruber1995,
  Guarino1998}.  Ontologies commonly distinguish between {\em classes}
and {\em individuals} within a domain of knowledge. A class is an
entity that can have {\em instances}, while individuals are entities
that cannot be instantiated. Examples of individuals include the
Eiffel tower or the 2009 Ironman Triathlon in Hawaii, while examples
of classes include {\em Tower} or {\em Triathlon}. The Eiffel tower
can be an instance of the class {\em Tower}, and the 2009 Ironman
Triathlon an instance of {\em Triathlon}.  The meaning of classes is
specified by stating what must be true of their instances.

@@

A first kind of abnormality of regulatory processes are {\em abnormal
  temporal distribution patterns} of a process. In these
abnormalities, the {\em way} in which processes of a particular kind
are temporally distributed is abnormal.  The most common abnormal
distribution pattern is an increased or decreased frequency, and we
use PATO's {\em frequency} class to define {\em Abnormal frequency of
  occurrence of $X$}.
% \begin{verbatim}
% has-phenotype some (has-part-some (participates-in some 
%   (regulates some X and has-quality some (frequency and towards some X))))
% \end{verbatim}
For example, an {\em Abnormal frequency of occurrence of apoptosis} is
defined as an abnormality of {\em Regulation of apoptosis} ({\tt
  GO:0042981}) with respect to the {\em frequency} ({\tt
  PATO:0000044}) of {\em Apoptosis} ({\tt GO:0006915}) occurrences.

A second kind of regulatory abnormalities are related to the {\em
  onset} of a process. With respect to a reference process, a
particular kind of process may be {\em delayed} ({\tt PATO:0000502})
or {\em premature} ({\tt PATO:0000694}). For example,

Further kinds include deviations from distribution patterns. For
example, a kind of process that is normally {\em rhythmic} can be
abnormal in that it is {\em arrhythmic}...
